---
title: "Hierarchical Clustering"
output:
  html_document:
    df_print: paged
---

## Introductory Example and Cluster analysis

The linkage criterium is the decision rule used to decide which of the 
observations is the closest to an already existing group. There are different 
linkage criteria (linkage methods):

- complete linkage : based on the maximum distance between two sets
- single linkage : based on the minimum distance between two sets
- average linkage : based on the average distance between two sets

The order in which the observations are grouped, which is determined by the 
linkage criterium (method), results in a hierarchy based on distance, hence, 
hierarchical clustering.

Hierarchical clustering can now be used to group the observations in a number 
of pre-defined clusters, i.e., k clusters.

```{r}
# Load packages
library(tidyverse)
library(dendextend)
```

```{r}
# Data frame with player positions
players <- data.frame(
  x = c(-1, -2, 8, 7, -12, -15, -13, 15, 21, 12, -25, 26),
  y = c(1, -3, 6, -8, 8, 0, -10, 16, 2, -15, 1, 0)
)
```

```{r}
# Distance matrix
dist_players <- dist(players, method = "euclidean")

# Perform the iterative linkage steps using the complete linkage method
hc_players <- hclust(dist_players, method = "complete")

# Extract two clusters
cluster_assignments <- cutree(hc_players, k = 2)

# Assign the cluster IDs to the players
clustered_players <- dplyr::mutate(players, cluster = cluster_assignments)
```

## Assessing Results of Cluster Analysis

Clustering analysis is in part a qualitative analysis. Thus, the necessary 
tools must be known to assess (or evaluate) the results of the cluster 
analysis.

```{r}
# Count the number of players in each cluster
dplyr::count(clustered_players, cluster)
```

```{r}
# Visualize the positions of the players
ggplot(clustered_players, aes(x, y, color = factor(cluster))) +
  geom_point()
```

Remember, the results of hierarchical cluster analysis are formed by the 
following choices:

- the method by which the distance is calculated, e.g. Euclidean
- the linkage method, e.g. complete linkage
- the number of pre-defined cluster, i.e., k clusters.

## Using the Dendrogram

The dendrogram is used to visualize how the observations within the data set 
are iteratively linked with one another. One specific characteristic of the 
dendrogram is that the branches encode the distance of the observations falling 
under that branch.

```{r}
# Create dendrogram
plot(hc_players)
```

Taking into account the linkage method, the dendrogram can be used to visualize 
the clustering of observations based on a distance cutoff value, i.e., a height 
cutoff value.

```{r}
# Create dendrogram object
dend_players <- as.dendrogram(hc_players)
plot(dend_players)
```

```{r}
# Cluster based on distance
dend_colored <- color_branches(dend_players, h = 35)
plot(dend_colored)
```

The cutree() function can also be used to create clusters not based on a number 
of pre-defined clusters, i.e., k clusters, but on a cutoff height.

```{r}
# Create cluster assignments based on cutoff height of 20
cluster_assignments_h20 <- cutree(hc_players, h = 20)
players_clustered_h20 <- mutate(players, cluster = cluster_assignments_h20)

# Visualize clustering
ggplot(players_clustered_h20, aes(x, y, color = factor(cluster))) +
  geom_point()
```

```{r}
# Create cluster assignments based on cutoff height of 40
cluster_assignments_h40 <- cutree(hc_players, h = 40)
players_clustered_h40 <- mutate(players, cluster = cluster_assignments_h40)

# Visualize clustering
ggplot(players_clustered_h40, aes(x, y, color = factor(cluster))) +
  geom_point()
```


















